---
title: "whitebox Tool Metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{whitebox Tool Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6 
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# R package dataset symbol names
wbttools <- NULL
wbttoolparameters <- NULL

data("wbttools", package = "whitebox")
```

# `whitebox` Tool Metadata

This vignette provides an introduction to the data sets included in the `whitebox` package. These data sets contain names, arguments and other metadata for tools available in WhiteboxTools.

### What version of WhiteboxTools are these data sets generated from?

These data sets are built to correspond to tool names available in the most recent version of WhiteboxTools and correspond to tools available in the R package.

 - Current version: **`r attr(wbttools, "version")`**

If you have a relatively recent version of WhiteboxTools the version should not be an issue, but the data sets are not dynamically generated from your whitebox installation.

## WhiteboxTools Tool Names and R Function Names

The first data set provides the relationship between tool names in WhiteboxTools and the corresponding exported function in the R package.

It also gives the WhiteboxTools Toolbox name and a brief description.

```{r}
data("wbttools", package = "whitebox")

str(wbttools)
```

The `wbttools` data set is a data.frame with `r nrow(wbttools)` tools and 4 variables

 - `"tool_name"` - WhiteboxTools tool name

 - `"toolbox_name"` - WhiteboxTools toolbox name

 - `"description"` - Brief description

 - `"function_name"` - R function name
 
```{r}
head(wbttools)
```

The R function naming style differs from the tool names in WhiteboxTools, but the core words are the same.

There are two steps to converting WhiteboxTools to R (if not using `wbttools`!)
 
 1. `CamelCase` tool names change to `snake_case` R function names
 
 2. All R function names have the prefix `wbt_` and add `()`
 
So, `StreamSlopeContinuous` becomes `wbt_stream_slope_continuous()` in R, for instance.
 
## WhiteboxTools Tool and Parameter Names

The second data set provides details about the options by tool name. 

```{r}
data("wbttoolparameters", package = "whitebox")
```

The `wbttoolparameters` data set is a _data.frame_ with `r nrow(wbttoolparameters)` parameters and `r ncol(wbttoolparameters)` variables:

 - `"function_name"` - R function name
 
 - `"tool_name"` - WhiteboxTools tool name
 
 - `"name"` - WhiteboxTools tool parameter name
 
 - `"flags"` - flags used to specify parameter on command line; comma-separated
 
 - `"description"` - parameter description
 
 - `"parameter_class"` - parameter type
 
 - `"parameter_detail"` - parameter details; character: data type followed by colon and more specifics, For OptionList possible values, comma-separated (if defined)

 - `"default_value"` - parameter default value, if any
 
 - `"optional"` - parameter "optional" flag; note that some combination of optional parameters may be required for certain conditions
 
 - `"argument_name"` - labels for selected subset of `"flags"` **used as R function argument names** for `wbt_` functions

 - `"is_input"` - logical. Classification of 'input' parameters
 
 - `"is_output"` - logical. Classification of 'output' parameters
 

```{r}
head(wbttoolparameters)
```

Several fields in this table such as `flags` and `parameter_type` are "flattened" relative to the nested `wbt_tool_parameters()` output.

The nested `parameter_type` from the JSON result is replaced with two variables in the data set: `parameter_class` and `parameter_details`

This parameter _data.frame_ is useful to construct your own functions with `wbt_run_tool()` or for inspecting the types of tools that can be run on particular data types.

```{r}
str(wbttoolparameters, max.level = 1)
```

You will find that both `tool_name` and `function_name` are present, so you can look up by whatever is convenient. 

The variable `argument_name` is processed to be the subset of `flags` that corresponds to arguments to R functions, which are denoted with `function_name`.

### Example: Finding Tools by Parameter Type

To find the tools that have an "OptionList" argument with parameter label "variant" (flag: `--variant=`) we can use `subset()`

```{r}
subset(wbttoolparameters, grepl("OptionList", parameter_class) & grepl("variant", argument_name))
```

Check out the code that builds the data in `/data-raw/` on [GitHub](https://github.com/giswqs/whiteboxR/tree/wbtDATA/data-raw) to see how these objects can be generated for any version of WhiteboxTools that supports JSON output for parameter metadata.

## Appendix: Tables of Function Names by Toolbox

The remainder of this vignette are tables of R function names and tool descriptions from `wbttoolparameters`, organized by WhiteboxTools toolbox name.

```{r, echo=FALSE, results='asis'}
# hide this, we add the parentheses for the docs
wbttoolsshow <- wbttools
wbttoolsshow$function_name <- paste0("`", wbttoolsshow$function_name, "()`")
wbttoolssplt <- split(wbttoolsshow[,colnames(wbttoolsshow) %in% c("function_name", "description")],
                      list(wbttoolsshow$toolbox_name))

if (requireNamespace("knitr")) {
  x <- sapply(names(wbttoolssplt), function(nm) {
    cat("###", nm, "\n\n")
    print(knitr::kable(
      wbttoolssplt[[nm]],
      caption = paste("Toolbox:", nm),
      row.names = FALSE
    ))
  })
}
```
