---
title: "whitebox examples"
author: "Qiusheng Wu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{whitebox examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## About WhiteboxTools

Load the whitebox R package.
```{r message=FALSE, warning=FALSE}
library(whitebox)
```

Prints the whitebox-tools help...a listing of available commands.
```{r}
print(wbt_help())
```

Prints the whitebox-tools license.
```{r}
print(wbt_license())
```
 
Prints the whitebox-tools version.
```{r}
print(wbt_version())
```

Prints the toolbox for a specific tool.
```{r eval=FALSE, message=FALSE}
print(wbt_toolbox())
```


List all available tools in whitebox-tools.
```{r eval=FALSE, message=FALSE}
print(wbt_list_tools())
```


Lists tools with 'lidar' in tool name or description.
```{r eval=FALSE, message=FALSE}
print(wbt_list_tools("lidar"))
```

Prints the help for a specific tool.
```{r}
print(wbt_tool_help("lidar_info"))
```

Retrieves the tool parameter descriptions for a specific tool.
```{r}
print(wbt_tool_parameters("slope"))
```

View the source code for a specific tool on the source code repository.
```{r}
print(wbt_view_code("breach_depressions"))
```


## How to run tools?
```{r eval=FALSE}
library(whitebox)
# Set input raster DEM file
dem <- system.file("extdata", "DEM.tif", package="whitebox")

# Run tools
wbt_feature_preserving_smoothing(dem, "./smoothed.tif", filter=9, verbose_mode = TRUE)
wbt_breach_depressions("./smoothed.tif", "./breached.tif")
wbt_d_inf_flow_accumulation(dem, "./flow_accum.tif")
```

